---
title: "stat602_hw2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 4.3

## First we read in the data set and remove the irrelavant columns.
```{r}
house <- read_excel("~/Downloads/AmesHousingData.xlsx")
house <- data.frame(Price = house$Price, Size = house$Size,
                    Fireplace = house$Fireplace,
                    Basementbath = house$`Bsmt Bath`, Land = house$Land, intersect = 1)
head(house)
```

## Now we need a matrix of all the possible regressor combinations.
```{r}
regMat <- expand.grid(c(TRUE,FALSE), c(TRUE,FALSE),
                      c(TRUE,FALSE), c(TRUE,FALSE))
regMat <- cbind(TRUE,regMat)

head(regMat)
```

## Then we construct a formula so that from each row of regMat we can determine the the regressor combinations.
```{r}
formu <- function(vec)
{
  vec <- as.matrix(vec)
  regressors <- c("intersect","Size", "Fireplace", "Basementbath", "Land")
  out <- as.formula(paste(c("Price ~ 0", regressors[vec]), collapse=" + "))
  
  return(out)
}

formu(regMat[1,])
```

## Cross-validation using **caret** package

We will repeat the 8-fold crossvalidation 10 times.
```{r, warning=FALSE}
ctrl <- trainControl(method = "repeatedcv", repeats = 10, number = 8)

RMSE <- numeric(16)
for(i in 1:length(RMSE)){
  model_caret <- train(formu(regMat[i,]), data = house, trControl = ctrl, method = "lm")
  RMSE[i] = model_caret$results[,2]
}
RMSE
order(RMSE)
```
We can see that the full model has the smallest mean squared prediction error.

# 4.4

